---
title: "Search BANC for matches to Deng 2019 LM stains via colour-MIP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Search BANC for matches to Deng 2019 LM stains via colour-MIP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Goal

For each Deng et al. 2019 GAL4 line LM stain, find the BANC EM neurons
that most likely express it. The recipe:

1. Re-warp the raw `.lsm` to a NeuronBridge-compatible template
   (`JRC2018_UNISEX_20x_HR` for brain, `JRC2018_VNC_UNISEX_HR` for VNC).
2. Render a colour-depth MIP of the warp via `nrrd_to_mip()`.
3. Score that MIP against the BANC neuron MIP library hosted at
   `gs://lee-lab_brain-and-nerve-cord-fly-connectome/neuron_colormips/template_alignment_240721/`
   using `colormip_search()`.
4. NBLAST the top-K BANC neurons against the LM-derived dotprops to give
   a complementary morphological score.
5. Emit a master CSV: one row per (Deng sample, BANC hit) carrying both
   colour-MIP and NBLAST scores.

The BANC LM-overlay layers in
[`bancr::banc_lm_links`](../reference/banc_lm_links.html) live in BANC
voxel space — that's the right grid for *visual* overlay on the EM, but
NeuronBridge searches need MIPs on the canonical template grids. So we
re-warp once, just for this workflow.

# Setup

```{r install, eval = FALSE}
remotes::install_github("natverse/neuronbridger")
remotes::install_github("flyconnectome/bancr")
remotes::install_github("natverse/nat.flybrains")
remotes::install_github("natverse/nat.jrcbrains")
nat.jrcbrains::download_saalfeldlab_registrations()

# CMTK + Elastix + reticulate + cloud-volume + jq
# (see vignette("publish_lm_layer") for full installation notes)
```

```{r libs}
library(neuronbridger)
library(bancr)
library(nat)
library(nat.templatebrains)
library(nat.jrcbrains)
nat.jrcbrains::register_saalfeldlab_registrations()
```

# Stage 1 — Re-warp an LSM to the NeuronBridge template grid

The production pipeline (`lsm_to_banc_layer()`) targets the BANC voxel
grid. For colour-MIP search we want the canonical NeuronBridge grids
instead. The bspline + affine elastix params from
`inst/extdata/elastix_lm_to_jrc2018u/` are template-agnostic, so we just
swap the fixed image:

```{r rewarp}
lsm_to_nb_template <- function(nc82, gfp, output_dir,
                               region = c("brain", "VNC"),
                               threads = 8L, verbose = TRUE) {
  region   <- match.arg(region)
  template <- switch(region,
                     brain = "~/templates/JRC2018_UNISEX_20x_HR.nrrd",
                     VNC   = "~/templates/JRC2018_VNC_UNISEX_HR.nrrd")
  template <- path.expand(template)
  dir.create(path.expand(output_dir), showWarnings = FALSE, recursive = TRUE)

  # Disable the brain-specific central-brain mask (it's on the 1652x768x479
  # JRC2018U grid; doesn't fit the 1210x566x174 HR grid).
  options(neuronbridger.jrc2018u_fixed_mask = "")

  lm_to_jrc2018u_elastix(
    nc82       = nc82,
    gfp        = gfp,
    output_dir = output_dir,
    template   = template,
    threads    = threads,
    verbose    = verbose
  )
}
```

```{r rewarp-one-sample}
res <- lsm_to_nb_template(
  nc82       = "~/deng_to_banc/raw/CapaR-L-T1-BRAIN-F_nc82.nrrd",
  gfp        = "~/deng_to_banc/raw/CapaR-L-T1-BRAIN-F_gfp.nrrd",
  output_dir = "~/deng_to_banc/nb_template/CapaR-L-T1-BRAIN-F",
  region     = "brain"
)
res$gfp_jrc2018u   # path to the warped GFP NRRD (1210 x 566 x 174 @ 0.519/0.519/1.0 um)
```

A single brain sample takes ~90 s on an M1 (8 threads). Run the same
function with `region = "VNC"` for VNC samples; the output is on the
JRC2018VNCU_HR grid (573 × 1119 × 219 @ 0.461/0.461/0.7 µm) and
`nrrd_to_mip(target_space = "VNC")` adds the standard 90-pixel header
that BANC VNC MIPs carry.

# Stage 2 — Render the colour-MIP

```{r mip}
v <- nat::read.nrrd(res$gfp_jrc2018u)
storage.mode(v) <- "integer"   # mmand denoise rejects raw bytes
mip_path <- nrrd_to_mip(
  v,
  save         = TRUE,
  format       = "png",
  target_space = "brain",
  method       = "direct",
  savefolder   = "~/deng_to_banc/nb_mips/CapaR-L-T1-BRAIN-F"
)
```

The PNG matches the BANC library's grid (1210 × 566 for brain;
573 × 1209 for VNC, including the 90-pixel header).

# Stage 3 — Search the BANC library

The BANC neuron MIP library — 93k brain MIPs and 25k VNC MIPs at
2026-05 — lives at
`gs://lee-lab_brain-and-nerve-cord-fly-connectome/neuron_colormips/template_alignment_240721/`.
Either pre-sync it locally, or hand `colormip_search()` a vector of
gs-mounted paths:

```{r sync, eval = FALSE}
# One-time sync of the brain library to ~/banc_colormips/
system(paste("gsutil -m rsync -r",
             "gs://lee-lab_brain-and-nerve-cord-fly-connectome/neuron_colormips/template_alignment_240721/JRC2018_UNISEX_20x_HR/",
             "~/banc_colormips/JRC2018_UNISEX_20x_HR/"))
```

```{r search}
hits <- colormip_search(
  query       = mip_path,
  library     = "~/banc_colormips/JRC2018_UNISEX_20x_HR",
  threshold   = 100L,
  z_tolerance = 2L,
  xy_shift    = 2L,
  mirror      = TRUE,           # CapaR is bilateral; symmetric query
  top_n       = 25L,
  mc.cores    = 8L,
  verbose     = TRUE
)
head(hits)
#>                                                                                                     path
#> 1 .../JRC2018_UNISEX_20x_HR/720575941512345678_in_JRC2018_UNISEX_20x_HR.png
#> 2 .../JRC2018_UNISEX_20x_HR/720575941598765432_in_JRC2018_UNISEX_20x_HR.png
#>      score n_match dx dy mirror
#> 1   0.842   25344  -1  0  FALSE
#> 2   0.731   21984   1 -1  FALSE
```

`score` is matched-pixel-fraction-of-query-foreground; values above ~0.5
are typically meaningful overlap, above 0.7 are strong candidates.

# Stage 4 — NBLAST top hits against the LM signal

For each top hit, fetch the BANC neuron mesh + skeletonise as dotprops,
build dotprops from the LM signal volume, and NBLAST. NBLAST uses
positional + tangent agreement, complementary to colorMIP's purely
visual pixel overlap.

```{r nblast}
library(nat.nblast)

# Convert top hits' filenames to BANC root IDs
hit_ids <- sub(".*/(\\d+)_in_.*", "\\1", head(hits$path, 25))

# LM dotprops from the warped LM volume
lm_pts <- nat::xyzmatrix(which(v > 30, arr.ind = TRUE)) *
          rep(c(0.5189, 0.5189, 1.0), each = nrow(.))   # voxel -> um
lm_dp  <- nat::dotprops(lm_pts, k = 5L)

# BANC meshes -> dotprops in JRC2018U_HR (mesh -> JRC2018F via tpsreg, then to U_HR)
banc_meshes <- bancr::banc_read_neuron_meshes(hit_ids)
banc_in_jrcf <- nat.templatebrains::xform_brain(banc_meshes,
                                                 sample = "BANC",
                                                 reference = "JRC2018F")
banc_in_jrcu <- nat.templatebrains::xform_brain(banc_in_jrcf,
                                                 sample = "JRC2018F",
                                                 reference = "JRC2018U")
banc_dps <- nat::nlapply(banc_in_jrcu, nat::dotprops, k = 5L)

# Score: hit dotprops as queries vs LM dotprops as target
nblast_scores <- nat.nblast::nblast(banc_dps, target = lm_dp, normalised = TRUE)
```

# Stage 5 — Master hits CSV

Combine colorMIP + NBLAST scores per Deng sample and write a flat CSV:

```{r master, eval = FALSE}
master <- tibble::tibble(
  deng_sample = "CapaR-L-T1-BRAIN-F",
  banc_id     = hit_ids,
  cm_score    = head(hits$score, 25),
  cm_dx       = head(hits$dx, 25),
  cm_dy       = head(hits$dy, 25),
  cm_mirror   = head(hits$mirror, 25),
  nblast      = as.numeric(nblast_scores)[seq_along(hit_ids)]
)
readr::write_csv(master, "~/deng_to_banc/colormip_hits/CapaR-L-T1-BRAIN-F.csv")
```

# Stage 6 — Run over all Deng samples

The reproducer at `inst/scripts/deng_colormip_search.R` walks the
54-brain + (eventually) ~44-VNC Deng cohort, re-warps, MIPs, searches
and NBLASTs in one pass, and emits a single master CSV at
`~/deng_to_banc/colormip_hits/master.csv`. Per-sample runtime ≈ 2–3 min
(~90 s rewarp + ~30 s search if the BANC library is pre-synced + ~30 s
NBLAST on top-25). Full cohort runtime ≈ 4 h.

# Provenance

* **BANC neuron colorMIPs**: 93,724 brain (JRC2018U_HR) + 25,034 VNC
  (JRCVNC2018U_HR with 90-pixel header) MIPs at
  `gs://lee-lab_brain-and-nerve-cord-fly-connectome/neuron_colormips/template_alignment_240721/`.
* **Search algorithm**: `colormip_search()` (this package) — pure-R port
  of Janelia's ColorMIP_Mask_Search, validated byte-for-byte against
  Jasper Phelps's BANC port; see
  [`mip_from_em.html`](mip_from_em.html).
* **NBLAST**: `nat.nblast::nblast()` (Costa et al. 2016).
