---
title: "Serving an LM volume as a Neuroglancer layer alongside BANC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serving an LM volume as a Neuroglancer layer alongside BANC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Goal

Take a light-microscopy stack registered to a fly template, resample
it into **BANC voxel coordinates** so it overlays correctly on the BANC
EM image, write it out in Neuroglancer
[**precomputed**](https://github.com/google/neuroglancer/tree/master/src/neuroglancer/datasource/precomputed)
format, and serve it as an extra layer in the canonical public BANC
scene.

> **A note on target spaces.** This vignette has a *different target
> space* than the colour-MIP vignettes:
>
> * The
>   [*colour-MIPs from a connectome neuron*](colormip_direct_vs_fiji.html)
>   and
>   [*colour-MIPs from a registered LM volume*](colormip_lm.html)
>   vignettes target **`JRC2018U_HR`** (or `JRC2018VNCU_HR` for VNC) —
>   the NeuronBridge ColorMIP search spaces. Their output is a 2-D
>   PNG that NeuronBridge can index.
>
> * **This vignette targets BANC voxel space** (`2400 × 924 × 789` at
>   `400 nm` for brain). A NG image layer only overlays on BANC EM if
>   it lives in the same voxel grid the EM does. Reaching that grid
>   requires the BANC team's published Elastix transform from
>   JRC2018F → BANC.
>
> So: a layer that's perfect for the NeuronBridge search workflow
> (JRC2018U_HR) won't overlay correctly on BANC EM in NG, and vice
> versa. Pipe each LM volume through whichever pipeline matches the
> downstream task.

The full pipeline has four stages:

| Stage | Tool | What it does |
|---|---|---|
| 1 | `nat::xform_brain()` (CMTK + `nat.h5reg`) | Source space (IS2 / JFRC2 / FCWB) → JRC2018F |
| 2 | `transformix -tp BANC_to_template.txt` | JRC2018F → BANC voxel coords (`2400 × 924 × 789 @ 400 nm`) |
| 3 | `nrrd_to_precomputed()` | BANC-aligned NRRD → Neuroglancer `precomputed` directory |
| 4 | `bancr::banc_lm_scene()` | Build a Spelunker scene with the new LM layer + the canonical public BANC scene |

# Setup

```{r install}
remotes::install_github("natverse/neuronbridger")
remotes::install_github("flyconnectome/bancr")        # banc_lm_scene + auth
remotes::install_github("natverse/nat.flybrains")
remotes::install_github("natverse/nat.jrcbrains")
nat.jrcbrains::download_saalfeldlab_registrations()   # ~ 10 GB; one-time

install.packages("reticulate")
reticulate::py_install("cloud-volume", pip = TRUE)    # the precomputed writer

# CMTK: pre-built MacOSX zip from https://www.nitrc.org/projects/cmtk/
# Java (for nat.h5reg): brew install openjdk  (or your platform's equivalent)
# Elastix 5.x: download from https://github.com/SuperElastix/elastix/releases/latest
```

```{r libs}
library(neuronbridger)
library(bancr)
library(nat)
library(nat.flybrains)
library(nat.templatebrains)
library(nat.jrcbrains)
nat.jrcbrains::register_saalfeldlab_registrations()
```

# Stage 1 — Source space → JRC2018F

The BANC team's Elastix chain expects input on the **JRC2018F** grid
(`1652 × 768 × 479` at 380 nm). For an IS2-space LM volume the
bridging chain is `IS2 → FCWB → JRC2018F` (CMTK then H5).

`nat.h5reg` exposes points-mode warping but not image-mode, so the
practical pipeline is:

1. Threshold + denoise the source stack (defaults: `nrrd_to_mip()`'s
   3 × 3 × 3 median + Triangle on 12-bit-trimmed data).
2. Take above-threshold voxel centres as 3-D points in source microns.
3. `xform_brain(points, sample = <source>, reference = "JRC2018F")`.
4. Voxelise the transformed points into the JRC2018F grid.

```{r is2-to-jrcf, eval = FALSE}
NRRD_IN <- "IS2_CapaR_no1_02_warp_m0g40c4e1e-1x16r3.nrrd"
v <- nat::read.nrrd(NRRD_IN)
voxdims_um <- diag(attr(v, "header")[["space directions"]])
vol <- as.integer(pmin(pmax(as.integer(v), 0L), 4095L) / 16L)
dim(vol) <- dim(v)

vol_med <- mmand::medianFilter(vol, mmand::shapeKernel(c(3, 3, 3), type = "box"))
thr <- neuronbridger:::colormip_triangle_threshold(vol_med)
fg_idx <- which(vol_med > thr, arr.ind = TRUE)
intens <- vol_med[vol_med > thr]
pts_is2 <- sweep(fg_idx - 1L, 2, voxdims_um, "*")

pts_jrcf <- nat.templatebrains::xform_brain(pts_is2,
                                            sample    = "IS2",
                                            reference = "JRC2018F")

# Voxelise into JRC2018F (1652 x 768 x 479 at 0.38 um isotropic),
# keeping max intensity per voxel.
ix <- as.integer(round(pts_jrcf[,1] / 0.38)) + 1L
iy <- as.integer(round(pts_jrcf[,2] / 0.38)) + 1L
iz <- as.integer(round(pts_jrcf[,3] / 0.38)) + 1L
keep <- !is.na(ix) & ix %in% 1:1652 & iy %in% 1:768 & iz %in% 1:479
ix <- ix[keep]; iy <- iy[keep]; iz <- iz[keep]; intens <- intens[keep]
vol_jrcf <- array(0L, dim = c(1652L, 768L, 479L))
lin <- ix + (iy - 1L) * 1652L + (iz - 1L) * 1652L * 768L
ord <- order(lin, intens); lin_s <- lin[ord]; intens_s <- intens[ord]
vol_jrcf[lin_s[!duplicated(lin_s, fromLast = TRUE)]] <-
  intens_s[!duplicated(lin_s, fromLast = TRUE)]
nat::write.nrrd(vol_jrcf, "CapaR_in_JRC2018F.nrrd")
```

# Stage 2 — JRC2018F → BANC voxel space (Elastix)

The BANC public bucket serves the JRC2018F template *pre-aligned to
BANC voxel coordinates* at
`gs://lee-lab_brain-and-nerve-cord-fly-connectome/templates/JRC2018F_aligned240721_to_BANC.ng/`,
produced by the Elastix transforms checked into the BANC repo at
[`fanc/transforms/transform_parameters/brain_240721/`](https://github.com/jasper-tms/the-BANC-fly-connectome/tree/main/fanc/transforms/transform_parameters/brain_240721).

*Confusingly, the file whose `Size` matches BANC (`2400 × 924 × 789
@ 400 nm`) is `BANC_to_template.txt` — that's the one that lands on
the BANC grid. Transformix chains the rest automatically.*

Run transformix with the JRC2018F input from Stage 1:

```{r elastix, eval = FALSE}
system(paste("transformix",
             "-in",  "CapaR_in_JRC2018F.nrrd",
             "-out", "./CapaR_BANC_xform_out",
             "-tp",  "brain_240721/BANC_to_template.txt"))
# Output: ./CapaR_BANC_xform_out/result.nrrd  (2400 x 924 x 789 at 400nm)
```

The float32 output has a few B-spline ringing pixels around the brain
margins — clip them and downcast to uint8 before precomputing:

```{r cleanup, eval = FALSE}
v <- nat::read.nrrd("CapaR_BANC_xform_out/result.nrrd")
v[v < 0.5] <- 0; v[v > 255] <- 255
v <- as.integer(round(v)); dim(v) <- c(2400L, 924L, 789L)
nat::write.nrrd(v, "CapaR_no1_02_aligned240721_to_BANC.nrrd",
                dtype = "byte", enc = "gzip")
```

# Stage 3 — Convert to Neuroglancer precomputed

`nrrd_to_precomputed()` (this package) reads an NRRD (or a 3-D R
array) and writes the on-disk format Neuroglancer expects: an `info`
JSON describing scales, chunk sizes and data type, plus
`<resolution>/<x_min-x_max>_<y_min-y_max>_<z_min-z_max>` chunks
(gzip'd raw by default).

```{r precomputed}
nrrd_to_precomputed(
  input      = "CapaR_no1_02_aligned240721_to_BANC.nrrd",
  output     = "/tmp/CapaR_BANC_pc",
  resolution = c(400, 400, 400),    # BANC voxel resolution
  data_type  = "uint8",
  encoding   = "raw",
  chunk_size = c(64L, 64L, 64L)     # match the public atlas chunk size
)
```

`inst/scripts/lm_capar_to_precomputed.R` ships a fuller reproducer
that down-samples 4× xy / 2× z and squashes 16-bit signal into 8-bit
to keep the demo precomputed dir small (~1.5 MB). For NeuronBridge
searches keep full resolution and `uint16` if you need the dynamic
range.

If you already have a Python pipeline, the
[`npimage`](https://github.com/jasper-tms/npimage) helper from the
BANC team wraps the same `cloud-volume` call as a one-liner:

```python
import npimage
arr = npimage.load("CapaR_in_JRC2018U.nrrd")
npimage.save(arr, "CapaR.ng", pixel_size=[519, 519, 1000])
# Source: https://github.com/jasper-tms/npimage/blob/main/npimage/imageio.py#L311-L366
```

The Python and R routes write byte-equivalent precomputed
directories; pick whichever fits your build environment.
**Multi-channel `.lsm` stacks** need a per-channel loop (or RGB
packing) on either side — neither helper auto-splits channels for
you.

# Stage 4 — Upload + assemble a BANC Neuroglancer scene

```{r upload}
# Lee-lab team members can mirror to the curated BANC LM bucket; everyone
# else needs their own writable, public-readable host.
system(paste("gsutil -m cp -r /tmp/CapaR_BANC_pc",
             "gs://lee-lab_brain-and-nerve-cord-fly-connectome/light_level/kondo_et_al_2020/CapaR_no1_02_aligned240721_to_BANC.ng/"))
```

> **Bucket caveat — important.**
> `gs://lee-lab_brain-and-nerve-cord-fly-connectome/` is a curated
> mirror maintained by the lee-lab BANC team and is **not
> public-write**. The `light_level/kondo_et_al_2020/` path above is
> the canonical home for the Kondo 2020 imports, but you'll need
> either (a) write access from the lee-lab maintainers, or (b) your
> own GCS / S3 / static-HTTP host that's **public-read** so
> Neuroglancer can fetch your chunks. Once your precomputed directory
> is reachable over HTTPS, pass its URL to `bancr::banc_lm_scene()`.

`bancr::banc_lm_scene()` then constructs a Neuroglancer state that
starts from the standard public BANC scene (BANC EM + segmentation +
region outlines + the JRC2018F atlas + imported FAFB / hemibrain /
MANC meshes) and appends your LM layer:

```{r scene}
u <- bancr::banc_lm_scene(
  lm_url     = paste0("gs://lee-lab_brain-and-nerve-cord-fly-connectome/",
                      "light_level/kondo_et_al_2020/",
                      "CapaR_no1_02_aligned240721_to_BANC.ng/CapaR_BANC_pc"),
  layer_name = "Kondo 2020 - CapaR (no1_02, aligned to BANC)",
  range      = c(1, 30),     # match the actual uint8 dynamic range
                              # (Elastix ringing was clipped at <0.5)
  opacity    = 0.55,          # default; matches the public atlas layer
  blend      = "additive",    # LM signal lights up where it overlaps EM
  volume_rendering = "on",    # required for 3-D rendering
  shorten    = TRUE,
  open       = TRUE
)
u
#> [1] "https://spelunker.cave-explorer.org/#!middleauth+https://global.daf-apis.com/nglstate/api/v1/5028046288453632"
```

[Pinned scene](https://spelunker.cave-explorer.org/#!middleauth+https://global.daf-apis.com/nglstate/api/v1/5028046288453632):
the Kondo 2020 CapaR stain on the canonical BANC scene. The CapaR
layer matches the public `JRC2018F atlas imported` layer's volume
rendering (`volumeRendering = "on"`, `depthSamples = 788`,
`opacity = 0.55`); only `shaderControls.normalized.range` differs —
`[29, 255]` for the bright template vs `[1, 30]` for the dim
post-Elastix LM. Tighten `range` for sparse stains; expand it for
brighter sources.

`shorten = TRUE` (default) POSTs the state via `bancr::banc_shorturl()`
(same helper `bancsee()` uses) and returns a
`spelunker.cave-explorer.org/...nglstate/api/v1/<id>` URL — needs a
CAVE token from `bancr::banc_set_token()`. `shorten = FALSE` skips
auth and inlines the state in a long fragment URL.

## Publish to `ng.banc.community/view/`

The public BANC viewer at
[`ng.banc.community/view/`](https://ng.banc.community/view/) and the
private CAVE-authenticated viewer at
[`ng.banc.community/`](https://ng.banc.community/) **don't accept
dynamically-POSTed states** the way Spelunker does. Instead they load
named states from `ngstate.banc.community/view/<state-name>`, which
redirects to JSON files committed under
[`the-BANC-fly-connectome/neuroglancer_states/view/`](https://github.com/jasper-tms/the-BANC-fly-connectome/tree/main/neuroglancer_states/view).

To publish your scene there:

1. Decode the spelunker URL into the underlying scene JSON and write
   it to disk:

   ```{r publish, eval = FALSE}
   sc <- fafbseg::ngl_decode_scene(u)
   writeLines(jsonlite::toJSON(sc, auto_unbox = TRUE,
                               null = "null", pretty = TRUE),
              "CapaR_no1_02_JRC2018U_HR.json")
   ```

2. PR `CapaR_no1_02_JRC2018U_HR.json` into
   `the-BANC-fly-connectome/neuroglancer_states/view/`.

3. Once merged, your scene loads from
   `https://ng.banc.community/view/#!CapaR_no1_02_JRC2018U_HR`.

For ad-hoc / dev sharing the spelunker URL is the practical pattern —
that's what `bancsee()` and `banc_lm_scene()` return by default.

# Self-contained sanity check (no IS2, no auth, no GCS)

The chunk below builds a tiny synthetic 3-D volume, writes it as
precomputed to a local directory, reads it back through `cloud-volume`
to confirm round-trip fidelity, and constructs a long-form BANC
Neuroglancer URL referencing it — entirely offline.

```{r demo, eval = TRUE}
library(neuronbridger)
library(reticulate)

set.seed(7)
v <- array(as.integer(runif(96 * 96 * 32, 0, 250)), dim = c(96L, 96L, 32L))

td <- tempfile()
out <- nrrd_to_precomputed(
  v,
  output     = td,
  resolution = c(519, 519, 1000),
  data_type  = "uint8",
  encoding   = "raw"
)

# Read back through cloud-volume; confirm we got our volume out unchanged.
np <- reticulate::import("numpy",       convert = TRUE)
cv <- reticulate::import("cloudvolume", convert = FALSE)
vol <- cv$CloudVolume(out, mip = 0L, fill_missing = TRUE)
back <- np$squeeze(np$asarray(vol[0:96, 0:96, 0:32]), axis = 3L)
identical(as.integer(back), as.integer(v))
```

```{r scene-demo, eval = TRUE}
# Build a BANC scene with the local layer (long fragment URL form; no
# auth needed). For a real sharable URL you'd upload the precomputed
# dir to a public bucket and pass that gs:// URL instead.
if (requireNamespace("bancr", quietly = TRUE)) {
  u <- bancr::banc_lm_scene(out, layer_name = "tiny synthetic",
                            shorten = FALSE)
  cat("URL prefix:", substring(u, 1, 100), "\n")
  cat("LM layer present:", grepl("synthetic|Synthetic", u, ignore.case = TRUE), "\n")
} else {
  message("Install bancr to assemble a BANC scene: ",
          "remotes::install_github('flyconnectome/bancr')")
}
```

# Provenance

* **BANC neuroglancer infrastructure**: the public layers (BANC EM,
  segmentation, JRC2018F atlas, region outlines, synapse cloud,
  imported FAFB / hemibrain / MANC meshes) are all served from
  `gs://lee-lab_brain-and-nerve-cord-fly-connectome/` — see
  `bancr::banc_scene()` for the canonical entry-point state.
* **Precomputed format**: Neuroglancer
  [precomputed spec](https://github.com/google/neuroglancer/tree/master/src/neuroglancer/datasource/precomputed),
  written here via the
  [`cloud-volume`](https://github.com/seung-lab/cloud-volume)
  Python library.
* **JRC2018F → BANC transform**: BANC connectome repo
  [`fanc/transforms/transform_parameters/brain_240721/`](https://github.com/jasper-tms/the-BANC-fly-connectome/tree/main/fanc/transforms/transform_parameters)
  (Elastix; TPS approximations for points are exposed as data in
  `bancr` — `banc_to_jrc2018f_tpsreg`).
