---
title: "Publish an LM volume as a Neuroglancer layer on BANC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publish an LM volume as a Neuroglancer layer on BANC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Goal

Take an LM stack registered to a fly template, resample it into
**BANC voxel coordinates** so it overlays correctly on BANC EM, write
it out as a Neuroglancer
[**precomputed**](https://github.com/google/neuroglancer/tree/master/src/neuroglancer/datasource/precomputed)
layer, and add it to the canonical public BANC scene.

> **Different target space than the MIP vignettes.** The MIP vignettes
> ([EM](mip_from_em.html), [LM](mip_from_lm.html)) target
> `JRC2018U_HR` / `JRC2018VNCU_HR` — the spaces NeuronBridge ColorMIP
> search expects. **This vignette targets BANC voxel space**
> (`2400 × 924 × 789` @ 400 nm for the brain), reached via the BANC
> team's Elastix transform from JRC2018F. The two outputs are not
> interchangeable: a `JRC2018U_HR` volume won't align on BANC EM, and
> a BANC-grid volume can't be NeuronBridge-indexed.

The full pipeline has four stages:

| Stage | Tool | What it does |
|---|---|---|
| 1 | `nat::xform_brain()` (CMTK + `nat.h5reg`) | Source space (IS2 / JFRC2 / FCWB) → JRC2018F |
| 2 | `transformix -tp BANC_to_template.txt` | JRC2018F → BANC voxel coords (`2400 × 924 × 789` @ 400 nm) |
| 3 | `nrrd_to_precomputed()` | BANC-aligned NRRD → Neuroglancer `precomputed` directory |
| 4 | `bancr::banc_lm_scene()` | Build a Spelunker scene with the new LM layer added to the canonical public BANC scene |

# Setup

```{r install}
remotes::install_github("natverse/neuronbridger")
remotes::install_github("flyconnectome/bancr")        # banc_lm_scene + auth
remotes::install_github("natverse/nat.flybrains")
remotes::install_github("natverse/nat.jrcbrains")
nat.jrcbrains::download_saalfeldlab_registrations()   # ~ 10 GB; one-time

install.packages("reticulate")
reticulate::py_install("cloud-volume", pip = TRUE)    # the precomputed writer

# CMTK: pre-built MacOSX zip from https://www.nitrc.org/projects/cmtk/
# Java (for nat.h5reg): brew install openjdk  (or your platform's equivalent)
# Elastix 5.x: download from https://github.com/SuperElastix/elastix/releases/latest
```

```{r libs}
library(neuronbridger)
library(bancr)
library(nat)
library(nat.flybrains)
library(nat.templatebrains)
library(nat.jrcbrains)
nat.jrcbrains::register_saalfeldlab_registrations()
```

# Stage 1 — Source space → JRC2018F

The BANC team's Elastix chain expects its input on the **JRC2018F**
grid (`1652 × 768 × 479` voxels at 380 nm). For an IS2-space LM
volume the bridging chain is `IS2 → FCWB → JRC2018F`: CMTK for the
first hop, `nat.h5reg` for the second.

`nat.h5reg` warps points but not whole image volumes, so we go via
points and re-voxelise:

1. Threshold + denoise the source stack (defaults: `nrrd_to_mip()`'s
   3 × 3 × 3 median + Triangle on 12-bit-trimmed data).
2. Treat the above-threshold voxel centres as a 3-D point cloud in
   source microns.
3. `xform_brain(points, sample = <source>, reference = "JRC2018F")`.
4. Voxelise the transformed points back onto the JRC2018F grid,
   keeping the maximum source intensity per output voxel.

```{r is2-to-jrcf, eval = FALSE}
NRRD_IN <- "IS2_CapaR_no1_02_warp_m0g40c4e1e-1x16r3.nrrd"
v <- nat::read.nrrd(NRRD_IN)
voxdims_um <- diag(attr(v, "header")[["space directions"]])
vol <- as.integer(pmin(pmax(as.integer(v), 0L), 4095L) / 16L)
dim(vol) <- dim(v)

vol_med <- mmand::medianFilter(vol, mmand::shapeKernel(c(3, 3, 3), type = "box"))
thr <- neuronbridger:::colormip_triangle_threshold(vol_med)
fg_idx <- which(vol_med > thr, arr.ind = TRUE)
intens <- vol_med[vol_med > thr]
pts_is2 <- sweep(fg_idx - 1L, 2, voxdims_um, "*")

pts_jrcf <- nat.templatebrains::xform_brain(pts_is2,
                                            sample    = "IS2",
                                            reference = "JRC2018F")

# Voxelise into JRC2018F (1652 x 768 x 479 at 0.38 um isotropic),
# keeping max intensity per voxel.
ix <- as.integer(round(pts_jrcf[,1] / 0.38)) + 1L
iy <- as.integer(round(pts_jrcf[,2] / 0.38)) + 1L
iz <- as.integer(round(pts_jrcf[,3] / 0.38)) + 1L
keep <- !is.na(ix) & ix %in% 1:1652 & iy %in% 1:768 & iz %in% 1:479
ix <- ix[keep]; iy <- iy[keep]; iz <- iz[keep]; intens <- intens[keep]
vol_jrcf <- array(0L, dim = c(1652L, 768L, 479L))
lin <- ix + (iy - 1L) * 1652L + (iz - 1L) * 1652L * 768L
ord <- order(lin, intens); lin_s <- lin[ord]; intens_s <- intens[ord]
vol_jrcf[lin_s[!duplicated(lin_s, fromLast = TRUE)]] <-
  intens_s[!duplicated(lin_s, fromLast = TRUE)]
nat::write.nrrd(vol_jrcf, "CapaR_in_JRC2018F.nrrd")
```

# Stage 2 — JRC2018F → BANC voxel space (Elastix)

The BANC public bucket serves the JRC2018F template already aligned
to BANC voxel coordinates at
`gs://lee-lab_brain-and-nerve-cord-fly-connectome/templates/JRC2018F_aligned240721_to_BANC.ng/`,
produced by the Elastix transforms checked into the BANC repo under
[`fanc/transforms/transform_parameters/brain_240721/`](https://github.com/jasper-tms/the-BANC-fly-connectome/tree/main/fanc/transforms/transform_parameters/brain_240721).
We use the same parameter chain to push our own JRC2018F-aligned LM
stack onto the BANC grid.

The naming is a little counter-intuitive: the parameter file whose
`Size` matches the BANC volume (`2400 × 924 × 789` @ 400 nm) is
`BANC_to_template.txt`, and that is the one to feed to transformix.
Transformix follows the chain back to BANC space automatically.

Run transformix with the JRC2018F volume from Stage 1 as input:

```{r elastix, eval = FALSE}
system(paste("transformix",
             "-in",  "CapaR_in_JRC2018F.nrrd",
             "-out", "./CapaR_BANC_xform_out",
             "-tp",  "brain_240721/BANC_to_template.txt"))
# Output: ./CapaR_BANC_xform_out/result.nrrd  (2400 x 924 x 789 at 400nm)
```

Transformix writes a float32 NRRD. B-spline interpolation produces a
small amount of ringing around the brain margins, with values just
below 0 or above 255. Clip those off and downcast to uint8 before
the precomputed step (otherwise an unsigned-integer cast wraps
negatives around to ~150 and looks like a uniform background haze):

```{r cleanup, eval = FALSE}
v <- nat::read.nrrd("CapaR_BANC_xform_out/result.nrrd")
v[v < 0.5] <- 0; v[v > 255] <- 255
v <- as.integer(round(v)); dim(v) <- c(2400L, 924L, 789L)
nat::write.nrrd(v, "CapaR_no1_02_aligned240721_to_BANC.nrrd",
                dtype = "byte", enc = "gzip")
```

# Stage 3 — Convert to Neuroglancer precomputed

`nrrd_to_precomputed()` (this package) reads an NRRD or a 3-D R array
and writes the on-disk layout Neuroglancer expects: an `info` JSON
describing scales, chunk sizes and data type, plus a flat directory
of chunks named `<resolution>/<x_min-x_max>_<y_min-y_max>_<z_min-z_max>`
(raw bytes by default; pass `compress = TRUE` to gzip them).

```{r precomputed}
nrrd_to_precomputed(
  input      = "CapaR_no1_02_aligned240721_to_BANC.nrrd",
  output     = "/tmp/CapaR_BANC_pc",
  resolution = c(400, 400, 400),    # BANC voxel resolution
  data_type  = "uint8",
  encoding   = "raw",
  chunk_size = c(64L, 64L, 64L)     # match the public atlas chunk size
)
```

`inst/scripts/lm_capar_to_precomputed.R` ships a fuller reproducer
that down-samples 4× in xy and 2× in z and squashes 16-bit signal
into 8-bit to keep the demo precomputed dir small (~1.5 MB). For
NeuronBridge searches keep the full resolution and use `uint16` if
the dynamic range matters.

If you already work in Python, the
[`npimage`](https://github.com/jasper-tms/npimage) helper from the
BANC team is a one-liner around the same `cloud-volume` call:

```python
import npimage
arr = npimage.load("CapaR_in_JRC2018U.nrrd")
npimage.save(arr, "CapaR.ng", pixel_size=[519, 519, 1000])
# Source: https://github.com/jasper-tms/npimage/blob/main/npimage/imageio.py#L311-L366
```

The Python and R routes write byte-equivalent precomputed
directories, so pick whichever fits your build environment.
**Multi-channel `.lsm` stacks** need a per-channel loop or RGB
packing on either side; neither helper splits channels for you.

# Stage 4 — Upload + assemble a BANC Neuroglancer scene

Neuroglancer fetches chunks over HTTPS, so the precomputed directory
needs a public-read host. Lee-lab BANC team members upload to the
curated mirror; everyone else uses their own bucket.

```{r upload}
system(paste("gsutil -m cp -r /tmp/CapaR_BANC_pc",
             "gs://lee-lab_brain-and-nerve-cord-fly-connectome/light_level/kondo_et_al_2020/CapaR_no1_02_aligned240721_to_BANC.ng/"))
```

> **Bucket caveat — important.**
> `gs://lee-lab_brain-and-nerve-cord-fly-connectome/` is a curated
> mirror run by the lee-lab BANC team and is **not public-write**.
> The `light_level/kondo_et_al_2020/` path above is the canonical
> home for the Kondo 2020 imports, but to write there you need either
> (a) write access granted by the lee-lab maintainers, or (b) your
> own GCS / S3 / static-HTTP host that is **public-read** so
> Neuroglancer can fetch the chunks. Once the precomputed directory
> is reachable over HTTPS, pass its URL to `bancr::banc_lm_scene()`.

> **One way to gzip, not two.** Either let
> `nrrd_to_precomputed(compress = TRUE)` write `.gz` chunks and
> upload them with plain `gsutil cp -r`, or write raw chunks
> (`compress = FALSE`) and upload with `gsutil cp -Z -r` so gsutil
> gzips on the fly and tags each blob with `Content-Encoding: gzip`.
> The two paths are equivalent; mixing them — `.gz` files uploaded
> with `-Z`, or raw files uploaded without it — yields a layer whose
> `info` is reachable but whose chunks all 404.

`bancr::banc_lm_scene()` builds a Neuroglancer state that starts from
the standard public BANC scene (BANC EM + segmentation + region
outlines + the JRC2018F atlas + imported FAFB / hemibrain / MANC
meshes) and appends your LM layer on top:

```{r scene}
u <- bancr::banc_lm_scene(
  lm_url     = paste0("gs://lee-lab_brain-and-nerve-cord-fly-connectome/",
                      "light_level/kondo_et_al_2020/",
                      "CapaR_no1_02_aligned240721_to_BANC.ng/CapaR_BANC_pc"),
  layer_name = "Kondo 2020 - CapaR (no1_02, aligned to BANC)",
  range      = c(1, 30),     # match the actual uint8 dynamic range
                              # (Elastix ringing was clipped at <0.5)
  opacity    = 0.55,          # default; matches the public atlas layer
  blend      = "additive",    # LM signal lights up where it overlaps EM
  volume_rendering = "on",    # required for 3-D rendering
  shorten    = TRUE,
  open       = TRUE
)
u
#> [1] "https://spelunker.cave-explorer.org/#!middleauth+https://global.daf-apis.com/nglstate/api/v1/5028046288453632"
```

[Pinned example scene](https://spelunker.cave-explorer.org/#!middleauth+https://global.daf-apis.com/nglstate/api/v1/5028046288453632):
the Kondo 2020 CapaR stain on the canonical BANC scene. The new LM
layer matches the public `JRC2018F atlas imported` layer's volume
rendering (`volumeRendering = "on"`, `depthSamples = 788`,
`opacity = 0.55`); only the `shaderControls.normalized.range` differs
— `[29, 255]` for the bright template vs `[1, 30]` for the dim
post-Elastix LM signal. Tighten `range` for sparse stains, expand it
for brighter sources.

`shorten = TRUE` (the default) POSTs the state via
`bancr::banc_shorturl()` — the same helper `bancsee()` uses — and
returns a `spelunker.cave-explorer.org/...nglstate/api/v1/<id>` URL.
This requires a CAVE token, set once with `bancr::banc_set_token()`.
Pass `shorten = FALSE` to skip the round-trip and inline the full
state in a long fragment URL instead.

# Self-contained sanity check (no IS2, no auth, no GCS)

The chunk below builds a tiny synthetic 3-D volume, writes it as
precomputed to a local directory, reads it back through `cloud-volume`
to confirm a clean round-trip, and constructs a long-form BANC
Neuroglancer URL referencing it — all offline. The chunk requires
`reticulate::py_install("cloud-volume", pip = TRUE)` and skips
automatically if the Python module isn't available.

```{r demo, eval = isTRUE(try(reticulate::py_module_available("cloudvolume"), silent = TRUE))}
library(neuronbridger)
library(reticulate)

set.seed(7)
v <- array(as.integer(runif(96 * 96 * 32, 0, 250)), dim = c(96L, 96L, 32L))

td <- tempfile()
out <- nrrd_to_precomputed(
  v,
  output     = td,
  resolution = c(519, 519, 1000),
  data_type  = "uint8",
  encoding   = "raw"
)

# Read back through cloud-volume; confirm we got our volume out unchanged.
np <- reticulate::import("numpy",       convert = TRUE)
cv <- reticulate::import("cloudvolume", convert = FALSE)
vol <- cv$CloudVolume(out, mip = 0L, fill_missing = TRUE)
back <- np$squeeze(np$asarray(vol[0:96, 0:96, 0:32]), axis = 3L)
identical(as.integer(back), as.integer(v))
```

```{r scene-demo, eval = isTRUE(try(reticulate::py_module_available("cloudvolume"), silent = TRUE)) && requireNamespace("bancr", quietly = TRUE)}
# Build a BANC scene with the local layer (long fragment URL form; no
# auth needed). For a real sharable URL you'd upload the precomputed
# dir to a public bucket and pass that gs:// URL instead.
u <- bancr::banc_lm_scene(out, layer_name = "tiny synthetic",
                          shorten = FALSE)
cat("URL prefix:", substring(u, 1, 100), "\n")
cat("LM layer present:", grepl("synthetic|Synthetic", u, ignore.case = TRUE), "\n")
```

# Provenance

* **BANC neuroglancer infrastructure**: the public layers (BANC EM,
  segmentation, JRC2018F atlas, region outlines, synapse cloud,
  imported FAFB / hemibrain / MANC meshes) are all served from
  `gs://lee-lab_brain-and-nerve-cord-fly-connectome/` — see
  `bancr::banc_scene()` for the canonical entry-point state.
* **Precomputed format**: Neuroglancer
  [precomputed spec](https://github.com/google/neuroglancer/tree/master/src/neuroglancer/datasource/precomputed),
  written here via the
  [`cloud-volume`](https://github.com/seung-lab/cloud-volume)
  Python library.
* **JRC2018F → BANC transform**: BANC connectome repo
  [`fanc/transforms/transform_parameters/brain_240721/`](https://github.com/jasper-tms/the-BANC-fly-connectome/tree/main/fanc/transforms/transform_parameters)
  (Elastix; TPS approximations for points are exposed as data in
  `bancr` — `banc_to_jrc2018f_tpsreg`).
